\invisiblesection{Appendix B: Glossary}

\thispagestyle{empty}

\vspace*{15em}

\begin{center}

\Huge
Appendix B

\end{center}

\cleardoublepage

The following glossary contains names and brief descriptions of individuals and entities referenced in the two volumes of this report.
It is not intended to be comprehensive and is intended only to assist a reader in the reading the rest of the report.

\subsection{Referenced Persons}

% \begin{longtablelarx}{\textwidth}{sb}
\begin{longtable}{ p{.35\textwidth} p{.65\textwidth} }

    \textbf{Agalarov, Aras} & Russian real-estate developer (owner of the Crocus Group); met Donald Trump in connection with the Miss Universe pageant and helped arrange the June~9, 2016 meeting at Trump Tower between Natalia Veselnitskaya and Trump Campaign officials. \\

    \textbf{Agalarov, Emin} & Performer, executive vice president of Crocus Group, and son of Aras Agalarov; helped arrange the June~9, 2016 meeting at Trump Tower between Natalia Veselnitskaya and Trump Campaign officials. \\

    \textbf{Akhmetov, Rinat} & Former member in the Ukrainian parliament who hired Paul Manafort to conduct work for Ukrainian political party, the Party of Regions. \\

    \textbf{Akhmetshin, Rinat} & U.S. lobbyist and associate of Natalia Veselnitskaya who attended the June~9, 2016 meeting at Trump Tower between Veselnitskaya and Trump Campaign officials. \\

    \textbf{Aslanovy, Dzheykhun (Jay)} & Head of U.S. department of the Internet Research Agency, which engaged in an ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election. \\

    \textbf{Assange, Julian} & Founder of WikiLeaks, which in 2016 posted on the internet documents stolen from entities and individuals affiliated with the Democratic Party. \\

    \textbf{Aven, Petr} & Chairman of the board of Alfa-Bank who attempted outreach to the Presidential Transition Team in connection with anticipated post-election sanctions. \\

    \textbf{Bannon, Stephen (Steve)} & White House chief strategist and senior counselor to President Trump (Jan.~2017--Aug.~2017); chief executive of the Trump Campaign. \\

    \textbf{Baranov, Andrey} & Director of investor relations at Russian state-owned oil company, Rosneft, and associate of Carter Page. \\

    \textbf{Berkowitz, Avi} & Assistant to Jared Kushner. \\

    \textbf{Boente, Dana} & Acting Attorney General (Jan.~2017--Feb.~2017); Acting Deputy Attorney General (Feb.~2017--Apr.~2017). \\

    \textbf{Bogacheva, Anna} & Internet Research Agency employee who worked on ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election; traveled to the United States under false pretenses in 2014. \\

    \textbf{Bossert, Thomas (Tom)} & Former homeland security advisor to the President who also served as a senior official on the Presidential Transition Team. \\

    \textbf{Boyarkin, Viktor} & Employee of Russian oligarch Oleg Deripaska. \\

    \textbf{Boyd, Charles} & Chairman of the board of directors at the Center for the National Interest, a U.S.-based think tank with operations in and connections to Russia. \\

    \textbf{Boyko, Yuriy} & Member of the Ukrainian political party Opposition Bloc and member of the Ukrainian parliament. \\

    \textbf{Brand, Rachel} & Associate Attorney General (May 2017--Feb.~2018). \\

    \textbf{Browder, William (Bill)} & Founder of Hermitage Capital Management who lobbied in favor of the Magnitsky Act, which imposed financial and travel sanctions on Russian officials. \\

    \textbf{Bulatov, Alexander} & Russian intelligence official who associated with Carter Page in 2008. \\

    \textbf{Burchik, Mikhail} & Executive director of the Internet Research Agency, which engaged in an ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election. \\

    \textbf{Burck, William} & Personal attorney to Don McGahn, White House Counsel. \\

    \textbf{Burnham, James} & Attorney in the White House Counsel's Office who attended January 2017 meetings between Sally Yates and Donald McGahn. \\

    \textbf{Burt, Richard} & Former U.S. ambassador who had done work [for] Alfa-Bank and was a board member of the Center for the National Interest. \\

    \textbf{Bystrov, Mikhail} & General director of the Internet Research Agency, which engaged in an ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election. \\

    \textbf{Calamari, Matt} & Chief operating officer for the Trump Organization. \\

    \textbf{Caputo, Michael} & Trump Campaign advisor. \\

    \textbf{Chaika, Yuri} & Prosecutor general of the Russian Federation who also maintained a relationship with Aras Agalarov. \\

    \textbf{Christie, Chris} & Former Governor of New Jersey. \\

    \textbf{Clapper, James} & Director of National Intelligence (Aug.~2010--Jan.~2017). \\

    \textbf{Clovis, Samuel Jr.} & Chief policy advisor and national co-chair of the Trump Campaign. \\

    \textbf{Coats, Dan} & Director of National Intelligence. \\

    \textbf{Cobb, Ty} & Special Counsel to the President (July 2017--May 2018). \\

    \textbf{Cohen, Michael} & Former vice president to the Trump Organization and special counsel to Donald Trump who spearheaded an effort to build a Trump-branded property in Moscow. He admitted to lying to Congress about the project. \\

    \textbf{Comey, James Jr.} & Director of the Federal Bureau of Investigation (Sept.~4, 2013--May~9, 2017). \\

    \textbf{Conway, Kellyanne} & Counselor to President Trump and manager of the Trump Campaign. \\

    \textbf{Corallo, Mark} & Spokesman for President Trump's personal legal team (June 2017--July 2017). \\

    \textbf{Corsi, Jerome} & Author and political commentator who formerly worked for WorldNetDaily and InfoWars. \blackout{Harm to Ongoing Investigation} \\

    \textbf{Costello, Robert} & Attorney who represented he had a close relationship with Rudolph Giuliani, the President's personal counsel. \\

    \textbf{Credico, Randolph (Randy)} & Radio talk show host who interviewed Julian Assange in 2016. \blackout{Harm to Ongoing Investigation} \\

    \textbf{Davis, Richard (Rick) Jr.} & Partner with Pegasus Sustainable Century Merchant Bank, business partner of Paul Manafort, and co-founder of the Davis Manafort lobbying firm. \\

    \textbf{Dearborn, Rick} & Former White House deputy chief of staff for policy who previously served as chief of staff to Senator Jeff Sessions. \\

    \textbf{Dempsey, Michael} & Office of Director of National Intelligence official who recalled discussions with Dan Coats after Coats's meeting with President Trump on March~22, 2017. \\

    \textbf{Denman, Diana} & Delegate to 2016 Republican National Convention who proposed a platform plank amendment that included armed support for Ukraine. \\

    \textbf{Deripaska, Oleg} & Russian businessman with ties to Vladimir Putin who hired Paul Manafort for consulting work between 2005 and~2009. \\

    \textbf{Dhillon, Uttam} & Attorney in the White House Counsel's Office (Jan.~2017--June 2018). \\

    \textbf{Dmitriev, Kirill} & Head of the Russian Direct Investment Fund (RDIF); met with Erik Prince in the Seychelles in January 2017 and, separately, drafted a U.S.--Russia reconciliation plan with Rick Gerson. \\

    \textbf{Donaldson, Annie} & Chief of staff to White House Counsel Donald McGahn (Jan.~2017--Dec.~2018). \\

    \textbf{Dvorkovich, Arkady} & Deputy prime minister of the Russian Federation and chairman of the board of directors of the New Economic School in Moscow. He met with Carter Page twice in 2016. \\

    \textbf{Dvoskin, Evgeney} & Executive of Genbank in Crimea and associate of Felix Sater. \\

    \textbf{Eisenberg, John} & Attorney in the White House Counsel's Office and legal counsel for the National Security Council. \\

    \textbf{Erchova, Lana (a/k/a Lana Alexander)} & Ex-wife of Dmitry Klokov who emailed Ivanka Trump to introduce Klokov to the Trump Campaign in the fall of 2015. \\

    \textbf{Fabrizio, Anthony (Tony)} & Partner at the research and consulting firm Fabrizio, Lee \& Associates. He was a pollster for the Trump Campaign and worked with Paul Manafort on Ukraine-related polling after the election. \\

    \textbf{Fishbein, Jason} & Attorney who performed worked for Julian Assange and also sent WikiLeaks a password for an unlaunched website \UseVerb{PutinTrumporg} on September~20, 2016. \\

    \textbf{Flynn, Michael G. (a/k/a Michael Flynn Jr.)} & Son of Michael T. Flynn, National Security Advisor (Jan.~20, 2017--Feb.~13, 2017). \\

    \textbf{Flynn, Michael T.} & National Security Advisor (Jan.~20, 2017--Feb.~13, 2017), Director of the Defense Intelligence Agency (July 2012--Aug.~7, 2014), and Trump Campaign advisor. He pleaded guilty to lying to the FBI about communications with Ambassador Sergey Kislyak in December 2016. \\

    \textbf{Foresman, Robert (Bob)} & Investment banker who sought meetings with the Trump Campaign in spring 2016 to discuss Russian foreign policy, and after the election met with Michael Flynn. \\

    \textbf{Futerfas, Alan} & Outside counsel for the Trump Organization and subsequently personal counsel for Donald Trump~Jr. \\

    \textbf{Garten, Alan} & General counsel of the Trump Organization. \\

    \textbf{Gates, Richard (Rick)~III} & Deputy campaign manager for Trump Campaign, Trump Inaugural Committee deputy chairman, and longtime employee of Paul Manafort. He pleaded guilty to conspiring to defraud the United States and violate U.S. laws, as well as making false statements to the~FBI\null. \\

    \textbf{Gerson, Richard (Rick)} & New York hedge fund manager and associate of Jared Kushner. During the transition period, he worked with Kirill Dmitriev on a proposal for reconciliation between the United States and Russia. \\

    \textbf{Gistaro, Edward} & Deputy Director of National Intelligence for Intelligence Integration. \\

    \textbf{Glassner, Michael} & Political director of the Trump Campaign who helped introduce George Papadopoulos to others in the Trump Campaign. \\

    \textbf{Goldstone, Robert} & Publicist for Emin Agalarov who contacted Donald Trump~Jr.\ to arrange the June~9, 2016 meeting at Trump Tower between Natalia Veselnitskaya and Trump Campaign officials. \\

    \textbf{Gordon, Jeffrey (J.D.)} & National security advisor to the Trump Campaign involved in changes to the Republican party platform and who communicated with Russian Ambassador Sergey Kislyak at the Republican National Convention. \\

    \textbf{Gorkov, Sergey} & Chairman of Vnesheconombank (VEB), a Russian state-owned bank, who met with Jared Kushner during the transition period. \\

    \textbf{Graff, Rhona} & Senior vice-president and executive assistant to Donald J. Trump at the Trump Organization. \\

    \textbf{\blackout{Harm to Ongoing Investigation}} & \blackout{Harm to Ongoing Investigation} \\

    \textbf{Hawker, Jonathan} & Public relations consultant at FTI Consulting; worked with Davis Manafort International LLC on public relations campaign in Ukraine. \\

    \textbf{Heilbrunn, Jacob} & Editor of the National Interest, the periodical that officially hosted candidate Trump's April 2016 foreign policy speech. \\

    \textbf{Hicks, Hope} & White House communications director (Aug.~2017--Mar.~2018) and press secretary for the Trump Campaign. \\

    \textbf{Holt, Lester} & NBC News anchor who interviewed President Trump on May~11, 2017. \\

    \textbf{Hunt, Jody} & Chief of staff to Attorney General Jeff Sessions (Feb.~2017--Oct.~2017). \\

    \textbf{Ivanov, Igor} & President of the Russian International Affairs Council and former Russian foreign minister. Ivan Timofeev told George Papadopoulos that Ivanov advised on arranging a ``Moscow visit'' for the Trump Campaign. \\

    \textbf{Ivanov, Sergei} & Special representative of Vladimir Putin, former Russian deputy prime minister, and former FSB deputy director. In January 2016, Michael Cohen emailed the Kremlin requesting to speak to Ivanov. \\

    \textbf{Kasowitz, Marc} & President Trump's personal counsel (May 2017--July 2017). \\

    \textbf{Katsyv, Denis} & Son of Peter Katsyv; owner of Russian company Prevezon Holdings~Ltd.\ and associate of Natalia Veselnitskaya. \\

    \textbf{Katsyv, Peter} & Russian businessman and father of Denis Katsyv. \\

    \textbf{\blackout{Harm to Ongoing Investigation}} & \blackout{Harm to Ongoing Investigation} \\

    \textbf{Kaveladze, Irakli (Ike)} & Vice president at Crocus Group and Aras Agalarov's deputy in the United States. He participated in the June~9, 2016 meeting at Trump Tower between Natalia Veselnitskaya and Trump Campaign officials. \\

    \textbf{Kaverzina, Irina} & Employee of the Internet Research Agency, which engaged in an ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election. \\

    \textbf{Kelly, John} & White House chief of staff (July 2017--Jan.~2019). \\

    \textbf{Khalilzad, Zalmay} & U.S. special representative to Afghanistan and former U.S. ambassador. He met with Senator Jeff Sessions during foreign policy dinners put together through the Center for the National Interest. \\

    \textbf{Kilimnik, Konstantin} & Russian-Ukrainian political consultant and long-time employee of Paul Manafort assessed by the FBI to have ties to Russian intelligence. \\

    \textbf{Kislyak, Sergey} & Former Russian ambassador to the United States and current Russian senator from Mordovia. \\

    \textbf{Klimentov, Denis} & Employee of the New Economic School who informed high-ranking Russian government officials of Carter Page's July 2016 visit to Moscow. \\

    \textbf{Klimentov, Dmitri} & Brother of Denis Klimentov who contacted Kremlin press secretary Dmitri Peskov about Carter Page's July 2016 visit to Moscow. \\

    \textbf{Klokov, Dmitry} & Executive for PJSC Federal Grid Company of Unified Energy System and former aide to Russia's minister of energy. He communicated with Michael Cohen about a possible meeting between Vladimir Putin and candidate Trump. \\

    \textbf{Kobyakov, Anton} & Advisor to Vladimir Putin and member of the Roscongress Foundation who invited candidate Trump to the St.~Petersburg International Economic Forum. \\

    \textbf{Krickovic, Andrej} & Professor at the Higher School of Economics who recommended that Carter Page give a July 2016 commencement address in Moscow. \\

    \textbf{Krylova, Aleksandra} & Internet Research Agency employee who worked on ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election; traveled to the United States under false pretenses in 2014. \\

    \textbf{Kushner, Jared} & President Trump's son-in-law and senior advisor to the President. \\

    \textbf{Kuznetsov, Sergey} & Russian government official at the Russian Embassy to the United States who transmitted Vladimir Putin's congratulations to President-Elect Trump for his electoral victory on November~9, 2016. \\

    \textbf{Landrum, Pete} & Advisor to Senator Jeff Sessions who attended the September 2016 meeting between Sessions and Russian Ambassador Sergey Kislyak. \\

    \textbf{Lavrov, Sergey} & Russian minister of foreign affairs and former permanent representative of Russia to the United Nations. \\

    \textbf{Ledeen, Barbara} & Senate staffer and associate of Michael Flynn who sought to obtain Hillary Clinton emails during the 2016 U.S. presidential campaign period. \\

    \textbf{Ledeen, Michael} & Member of the Presidential Transition Team who advised on foreign policy and national security matters. \\

    \textbf{Ledgett, Richard} & Deputy director of the National Security Agency (Jan.~2014--Apr.~2017); present when President Trump called Michael Rogers on March~26, 2017. \\

    \textbf{Lewandowski, Corey} & Campaign manager for the Trump Campaign (Jan.~2015--June 2016). \\

    \textbf{Luff, Sandra} & Legislative director for Senator Jeff Sessions; attended a September 2016 meeting between Sessions and Russian Ambassador Sergey Kislyak. \\

    \textbf{Lyovochkin, Serhiy} & Member of Ukrainian parliament and member of Ukrainian political party, Opposition Bloc Party. \\

    \textbf{Magnitsky, Sergei} & Russian tax specialist who alleged Russian government corruption and died in Russian police custody in 2009. His death prompted passage of the Magnitsky Act, which imposed financial and travel sanctions on Russian officials. \\

    \textbf{Malloch, Theodore (Ted)} & Chief executive officer of Global Fiduciary Governance and the Roosevelt Group. He was a London-based associate of Jerome Corsi. \\

    \textbf{Manafort, Paul Jr.} & Trump campaign member (March 2016--Aug.~2016) and chairman and chief strategist (May 2016--Aug.~2016). \\

    \textbf{Mashburn, John} & Trump administration official and former policy director to the Trump Campaign. \\

    \textbf{McCabe, Andrew} & Acting director of the FBI (May 2017--Aug.~2017); deputy director of the FBI (Feb.~2016--Jan.~2018). \\

    \textbf{McCord, Mary} & Acting Assistant Attorney General (Oct.~2016--May~2017). \\

    \textbf{McFarland, Kathleen (K.T.)} & Deputy White House National Security Advisor (Jan.~2017--May 2017). \\

    \textbf{McGahn, Donald (Don)} & White House Counsel (Jan.~2017--Oct.~2018). \\

    \textbf{Medvedev, Dmitry} & Prime Minister of Russia. \\

    \textbf{Melnik, Yuriy} & Spokesperson for the Russian Embassy in Washington, D.C., who connected with George Papadopoulos on social media. \\

    \textbf{Mifsud, Joseph} & Maltese national and former London-based professor who, immediately after returning from Moscow in April 2016, told George Papadopoulos that the Russians had ``dirt'' in the form of thousands of Clinton emails. \\

    \textbf{Miller, Matt} & Trump Campaign staff member who was present at the meeting of the National Security and Defense Platform Subcommittee in July 2016. \\

    \textbf{Miller, Stephen} & Senior advisor to the President.  \\

    \textbf{Millian, Sergei} & Founder of the Russian American Chamber of Commerce who met with George Papadopoulos during the campaign. \\

    \textbf{Mnuchin, Steven} & Secretary of the Treasury. \\

    \textbf{\blackout{Harm to Ongoing Investigation}} & \blackout{Harm to Ongoing Investigation} \\

    \textbf{M{\"u}ller-Maguhn, Andrew} & Member of hacker association Chaos Computer Club and associate of Julian Assange, founder of WikiLeaks. \\

    \textbf{Nader, George} & Advisor to the United Arab Emirates's Crown Prince who arranged a meeting between Kirill Dmitriev and Erik Prince during the transition period. \\

    \textbf{Netyksho, Viktor} & Russian military officer in command of a unit involved in Russian hack-and-release operations to interfere in the 2016 U.S. presidential election. \\

    \textbf{Oganov, Georgiy} & Advisor to Oleg Deripaska and a board member of investment company Basic Element. He met with Paul Manafort in Spain in early 2017. \\

    \textbf{Oknyansky, Henry (a/k/a Henry Greenberg)} & Florida-based Russian individual who claimed to have derogatory information pertaining to Hillary Clinton. He met with Roger Stone in May~2016. \\

    \textbf{Page, Carter} & Foreign policy advisor to the Trump Campaign who advocated pro-Russian views and made July 2016 and December 2016 visits to Moscow. \\

    \textbf{Papadopoulos, George} & Foreign policy advisor to the Trump Campaign who received information from Joseph Mifsud that Russians had ``dirt'' in the form of thousands of Clinton emails. He pleaded guilty to lying to the FBI about his contact with Mifsud. \\

    \textbf{Parscale, Bradley} & Digital media director for the 2016 Trump Campaign. \\

    \textbf{Patten, William (Sam) Jr.} & Lobbyist and business partner of Konstantin Kilimnik. \\

    \textbf{Peskov, Dmitry} & Deputy chief of staff of and press secretary for the Russian presidential administration. \\

    \textbf{Phares, Walid} & Foreign policy advisor to the Trump Campaign and co-secretary general of the Transatlantic Parliamentary Group on Counterterrorism (TAG). \\

    \textbf{Pinedo, Richard} & U.S. person who pleaded guilty to a single-count information of identity fraud. \\

    \textbf{Podesta, John Jr.} & Clinton campaign chairman whose email account was hacked by the GRU\null. WikiLeaks released his stolen emails during the 2016 campaign. \\

    \textbf{Podobnyy, Victor} & Russian intelligence officer who interacted with Carter Page while operating inside the United States; later charged in 2015 with conspiring to act as an unregistered agent of Russia. \\

    \textbf{Poliakova, Elena} & Personal assistant to Dmitry Peskov who responded to Michael Cohen's outreach about the Trump Tower Moscow project in January 2016. \\

    \textbf{Polonskaya, Olga} & Russian national introduced to George Papadopoulos by Joseph Mifsud as an individual with connections to Vladimir Putin. \\

    \textbf{Pompeo, Michael} & U.S. Secretary of State; director of the Central Intelligence Agency (Jan.~2017--Apr.~2018). \\

    \textbf{Porter, Robert} & White House staff secretary (Jan.~2017--Feb.~2018). \\

    \textbf{Priebus, Reince} & White House chief of staff (Jan.~2017--July 2017); chair of the Republican National Committee (Jan.~2011--Jan.~2017). \\

    \textbf{Prigozhin, Yevgeniy} & Head of Russian companies Concord Catering and Concord Management and Consulting; supported and financed the Internet Research Agency, which engaged in an ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election. \\

    \textbf{Prikhodko, Sergei} & First deputy head of the Russian Government Office and former Russian deputy prime minister. In January 2016, he invited candidate Trump to the St.~Petersburg International Economic Forum. \\

    \textbf{Prince, Erik} & Businessman and Trump Campaign supporter who met with Presidential Transition Team officials after the election and traveled to the Seychelles to meet with Kirill Dmitriev in January 2017. \\

    \textbf{Raffel, Josh} & White House communications advisor (Apr.~2017--Feb.~2018). \\

    \textbf{Rasin, Alexei} & Ukrainian associate of Henry Oknyansky who claimed to possess derogatory information regarding Hillary Clinton. \\

    \textbf{Rogers, Michael} & Director of the National Security Agency (Apr.~2014--May 2018). \\

    \textbf{Rosenstein, Rod} & Deputy Attorney General (Apr.~2017--present); Acting Attorney General for the Russian election interference investigation (May 2017--Nov.~2018). \\

    \textbf{Rozov, Andrei} & Chairman of I.C. Expert Investment Company, a Russian real-estate development corporation that signed a letter of intent for the Trump Tower Moscow project in 2015. \\

    \textbf{Rtskhiladze, Giorgi} & Executive of the Silk Road Transatlantic Alliance, LLC who communicated with Cohen about a Trump Tower Moscow proposal. \\

    \textbf{Ruddy, Christopher} & Chief executive of Newsmax Media and associate of President Trump. \\

    \textbf{Rybicki, James} & FBI chief of staff (May 2015--Feb.~2018). \\

    \textbf{Samochornov, Anatoli} & Translator who worked with Natalia Veselnitskaya and attended a June~9, 2016 meeting at Trump Tower between Veselnitskaya and Trump Campaign officials. \\

    \textbf{Sanders, Sarah Huckabee} & White House press secretary (July 2017--present). \\

    \textbf{Sater, Felix} & Real-estate advisor who worked with Michael Cohen to pursue a Trump Tower Moscow project. \\

    \textbf{Saunders, Paul J.} & Executive with the Center for the National Interest who worked on outlines and logistics of candidate Trump's April 2016 foreign policy speech. \\

    \textbf{Sechin, Igor} & Executive chairman of Rosneft, a Russian state[-]owned oil company. \\

    \textbf{Sessions, Jefferson~III (Jeff)} & Attorney General (Feb.~2017--Nov.~2018); U.S. Senator (Jan.~1997--Feb.~2017); head of the Trump Campaign's foreign policy advisory team. \\

    \textbf{Shoygu, Sergey} & Russian Minister of Defense. \\

    \textbf{Simes, Dimitri} & President and chief executive officer of the Center for the National Interest. \\

    \textbf{Smith, Peter} & Investment banker active in Republican politics who sought to obtain Hillary Clinton emails during the 2016 U.S. presidential campaign period. \\

    \textbf{Spicer, Sean} & White House press secretary and communications director (Jan.~2017--July 2017). \\

    \textbf{Stone, Roger} & Advisor to the Trump Campaign \blackout{Harm to Ongoing Investigation} \\

    \textbf{Tillerson, Rex} & U.S. Secretary of State (Feb.~2017--Mar.~2018). \\

    \textbf{Timofeev, Ivan} & Director of programs at the Russian International Affairs Council and program director of the Valdai Discussion Club who communicated in 2016 with George Papadopoulos, attempting to arrange a meeting between the Russian government and the Trump Campaign. \\

    \textbf{Trump, Donald Jr.} & President Trump's son; trustee and executive vice president of the Trump Organization; helped arrange and attended the June~9, 2016 meeting at Trump Tower between Natalia Veselnitskaya and Trump Campaign officials. \\

    \textbf{Trump, Eric} & President Trump's son; trustee and executive vice president of the Trump Organization. \\

    \textbf{Trump, Ivanka} & President Trump's daughter; advisor to the President and former executive vice president of the Trump Organization. \\

    \textbf{Ushakov, Yuri Viktorovich} & Aide to Vladimir Putin and former Russian ambassador to the United States; identified to the Presidential Transition Team as the proposed channel to the Russian government. \\

    \textbf{Vaino, Anton} & Chief of staff to Russian president Vladimir Putin. \\

    \textbf{Van der Zwaan, Alexander} & Former attorney at Skadden, Arps, Slate, Meagher \& Flom, LLP; worked with Paul Manafort and Rick Gates. \\

    \textbf{Vargas, Catherine} & Executive assistant to Jared Kushner. \\

    \textbf{Vasilchenko, Gleb} & Internet Research Agency employee who engaged in an ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election. \\

    \textbf{Veselnitskaya, Natalia} & Russian attorney who advocated for the repeal of the Magnitsky Act and was the principal speaker at the June~9, 2016 meeting at Trump Tower with Trump Campaign officials. \\

    \textbf{Weber, Shlomo} & Rector of the New Economic School (NES) in Moscow who invited Carter Page to speak at NES commencement in July 2016. \\

    \textbf{Yanukovych, Viktor} & Former president of Ukraine who had worked with Paul Manafort. \\

    \textbf{Yates, Sally} & Acting Attorney General (Jan.~20, 2017--Jan.~30, 2017); Deputy Attorney General (Jan.~10, 2015--Jan.~30, 2017). \\

    \textbf{Yatsenko, Sergey} & Deputy chief financial officer of Gazprom, a Russian state-owned energy company, and associate of Carter Page. \\

    \textbf{Zakharova, Maria} & Director of the Russian Ministry of Foreign Affair's Information and Press Department who received notification of Carter Page's speech in July 2016 from Denis Klimentov. \\

    \textbf{Zayed al Nahyan, Mohammed bin} & Crown Prince of Abu Dhabi and deputy supreme commander of the United Arab Emirates (UAE) armed forces.

\end{longtable}

\subsection{Entities and Organizations}

\begin{longtable}{ p{.35\textwidth} p{.65\textwidth} }

    \textbf{Alfa-Bank} & Russia's largest commercial bank, which is headed by Petr Aven. \\

    \textbf{Center for the National Interest (CNI)} & U.S.-based think tank with expertise in and connections to Russia. CNI's publication, the National Interest, hosted candidate Trump's foreign policy speech in April 2016. \\

    \textbf{Concord} & Umbrella term for Concord Management and Consulting, LLC and Concord Catering, which are Russian companies controlled by Yevgeniy Prigozhin. \\

    \textbf{Crocus Group or Crocus International} & A Russian real-estate and property development company that, in 2013, hosted the Miss Universe Pageant, and from 2013 through 2014, worked with the Trump Organization on a Trump Moscow project. \\

    \textbf{DCLeaks} & Fictitious online persona operated by the GRU that released stolen documents during the 2016 U.S. presidential campaign period. \\

    \textbf{Democratic Congressional Campaign Committee} & Political committee working to elect Democrats to the House of Representatives; hacked by the GRU in April 2016. \\

    \textbf{Democratic National Committee} & Formal governing body for the Democratic Party; hacked by the GRU in April 2016. \\

    \textbf{Duma} & Lower House of the national legislature of the Russian Federation. \\

    \textbf{Gazprom} & Russian oil and gas company majority-owned by the Russian government. \\

    \textbf{Global Energy Capital, LLC} & Investment and management firm founded by Carter Page. \\

    \textbf{Global Partners in Diplomacy} & Event hosted in partnership with the U.S. Department of State and the Republican National Convention. In 2016, Jeff Sessions and J.D. Gordon delivered speeches at the event and interacted with Russian Ambassador Sergey Kislyak. \\

    \textbf{Guccifer 2.0} & Fictitious online persona operated by the GRU that released stolen documents during the 2016 U.S. presidential campaign period. \\

    \textbf{I.C. Expert Investment Company} & Russian real-estate and development corporation that signed a letter of intent with a Trump Organization subsidiary to develop a Trump Moscow property. \\

    \textbf{Internet Research Agency (IRA)} & Russian entity based in Saint Petersburg and funded by Concord that engaged in an ``active measures'' social media campaign to interfere in the 2016 U.S. presidential election. \\

    \textbf{KLS Research LLC} & Business established by an associate of and at the direction of Peter Smith to further Smith's search for Hillary Clinton emails. \\

    \textbf{Kremlin} & Official residence of the president of the Russian Federation; it is used colloquially to refer to the office of the president or the Russian government. \\

    \textbf{LetterOne} & Company that includes Petr Aven and Richard Burt as board members. During a board meeting in December 2016, Aven asked for Burt's help to make contact with the Presidential Transition Team. \\

    \textbf{Link Campus University} & University in Rome, Italy, where George Papadopoulos was introduced to Joseph Mifsud. \\

    \textbf{London Centre of International Law Practice (LCILP)} & International law advisory organization in London that employed Joseph Mifsud and George Papadopoulos. \\

    \textbf{Main Intelligence Directorate of the General Staff (GRU)} & Russian Federation's military intelligence agency. \\

    \textbf{New Economic School in Moscow (NES)} & Moscow-based school that invited Carter Page to speak at its July 2016 commencement ceremony. \\

    \textbf{Opposition Bloc} & Ukrainian political party that incorporated members of the defunct Party of Regions. \\

    \textbf{Party of Regions} & Ukrainian political party of former President Yanukovych. It was generally understood to align with Russian policies. \\

    \textbf{Pericles Emerging Market Partners LLP} & Company registered in the Cayman Islands by Paul Manafort and his business partner Rick Davis. Oleg Deripaska invested in the fund. \\

    \textbf{Prevezon Holdings Ltd.} & Russian company that was a defendant in a U.S. civil action alleging the laundering of proceeds from fraud exposed by Sergei Magnitsky. \\

    \textbf{Roscongress Foundation} & Russian entity that organized the St.~Petersburg International Economic Forum. \\

    \textbf{Rosneft} & Russian state-owned oil and energy company. \\

    \textbf{Russian Direct Investment Fund} & Sovereign wealth fund established by the Russian Government in 2011 and headed by Kirill Dmitriev. \\

    \textbf{Russian International Affairs Council} & Russia-based nonprofit established by Russian government decree. It is associated with the Ministry of Foreign Affairs, and its members include Ivan Timofeev, Dmitry Peskov, and Petr Aven. \\

    \textbf{Silk Road Group} & Privately held investment company that entered into a licensing agreement to build a Trump-branded hotel in Georgia. \\

    \textbf{St.~Petersburg International Economic Forum} & Annual event held in Russia and attended by prominent Russian politicians and businessmen. \\

    \textbf{Tatneft} & Russian energy company. \\

    \textbf{Transatlantic Parliamentary Group on Counterterrorism} & European group that sponsored a summit between European Parliament lawmakers and U.S. persons. George Papadopoulos, Sam Clovis, and Walid Phares attended the TAG summit in July 2016. \\

    \textbf{Unit 26165 (GRU)} & GRU military cyber unit dedicated to targeting military, political, governmental, and non-governmental organizations outside of Russia. It engaged in computer intrusions of U.S. persons and organizations, as well as the subsequent release of the stolen data, in order to interfere in the 2016 U.S. presidential election. \\

    \textbf{Unit 74455 (GRU)} & GRU military unit with multiple departments that engaged in cyber operations. It engaged in computer intrusions of U.S. persons and organizations, as well as the subsequent release of the stolen data, in order to interfere in the 2016 U.S. presidential election. \\

    \textbf{Valdai Discussion Club} & Group that holds a conference attended by Russian government officials, including President Putin. \\

    \textbf{WikiLeaks} & Organization founded by Julian Assange that posts information online, including data stolen from private, corporate, and U.S. Government entities. Released data stolen by the GRU during the 2016 U.S. presidential election.

\end{longtable}

\subsection{Index of Acronyms}

\begin{longtable}{ p{.35\textwidth} p{.65\textwidth} }

    \textbf{CNI} & Center for the National Interest \\

    \textbf{DCCC} & Democratic Congressional Campaign Committee \\

    \textbf{DNC} & Democratic National Committee \\

    \textbf{FBI} & Federal Bureau of Investigation \\

    \textbf{FSB} & Russian Federal Security Service \\

    \textbf{GEC} & Global Energy Capital, LLC \\

    \textbf{GRU} & Russian Federation's Main Intelligence Directorate of the General Staff \\

    \textbf{HPSCI} & U.S. House of Representatives Permanent Select Committee on Intelligence \\

    \textbf{HRC} & Hillary Rodham Clinton \\

    \textbf{TRA} & Internet Research Agency \\

    \textbf{LCILP} & London Centre of International Law Practice \\

    \textbf{NATO} & North Atlantic Treaty Organization \\

    \textbf{NES} & New Economic School \\

    \textbf{NSA} & National Security Agency \\

    \textbf{ODNI} & Office of the Director of National Intelligence \\

    \textbf{PTT} & Presidential Transition Team \\

    \textbf{RDIF} & Russian Direct Investment Fund \\

    \textbf{RIAC} & Russian International Affairs Council \\

    \textbf{SBOE} & State boards of elections \\

    \textbf{SCO} & Special Counsel's Office \\

    \textbf{SIC} & U.S. Senate Judiciary Committee \\

    \textbf{SSCI} & U.S. Senate Select Committee on Intelligence \\

    \textbf{TAG} & Transatlantic Parliamentary Group on Counterterrorism \\

    \textbf{VEB} & Vnesheconombank

\end{longtable}

\cleardoublepage
